The Cost of Exclusion: Political Discrimination and Trust Deficits in the European Union

### 2025-07-11 article/main.tex
Bibliography added

While Europe has historically exhibited lower levels of affective polarization and declining partisan identification, recent studies suggest the presence of partisan hostility and negative feelings driven by factors such as ideological distance, perceived party threats, and regional tensions. Yet, evidence on longitudinal trends is still inconclusive. Scholars continue to emphasize the need for improved measurements and long-term data to determine whether affective polarization is indeed increasing across European democracies (Wagner, 2024).

Importantly, the rise of far-right populist parties has introduced a distinct affective dimension to European politics, one marked by widespread hostility toward political out-groups. This emerging divide signals the presence of a particularly polarized and emotionally charged layer of political competition (Reiljan, 2019; 2025).

Amid this environment, institutional trust is often viewed as a background condition for democratic stability. However, trust plays a more active and central role—it serves as a fundamental mechanism through which political systems gain legitimacy and foster public compliance. A robust body of literature supports the idea that institutional trust enhances democratic resilience, encouraging civic engagement and reinforcing institutional performance (Marien, 2011; Devine and Valgarosson, 2024).

@@

Institutional trust is often treated as a background condition for democratic stability. Yet, rather than serving as a passive backdrop, it functions as a central mechanism through which political systems earn legitimacy and secure compliance. Democratic institutions rely not only on performance or coercion but on public belief in the fairness of rules, the impartiality of procedures, and the integrity of leaders (Levi \& Stoker, 2000; Citrin \& Stoker, 2018).

Trust in institutions reflects more than evaluations of policy outcomes—it signals deeper commitments to institutional principles and the constitutional order (Easton, 1965). Building on this foundation, scholars have shown that institutional trust reinforces civic engagement and perceptions of procedural fairness (Tyler, 2006). More recent work highlights its crucial role in promoting democratic resilience, particularly during crises or episodes of institutional contestation (Zmerli \& Van der Meer, 2017; Devine \& Valgarosson, 2024).

Empirical evidence supports the view that trust strengthens democracy. For instance, Marien (2011) finds that institutional trust enables political participation, especially among less resourceful citizens. Other studies show that trust increases when institutions are seen as inclusive and responsive: Mann et al. (2022) report higher trust among sexual minorities in countries with equal rights legislation; Wilkes and Wu (2019) and Tyrberg (2023) observe similar patterns among ethnic minorities and immigrants.

@@

Supporting this view, Levy and Sacks (2009) show that perceived government unresponsiveness among disadvantaged groups is associated with lower trust and political efficacy. Anderson and Tverdova (2003) and Oskooii (2024) similarly find that perceived injustice, whether interpersonal or institutional, generates cognitive and emotional responses that undermine confidence in state institutions.

While this framework has typically focused on ethnic or cultural exclusion (Wilkes, 2018; Tyrberg, 2023), political exclusion—based on one's beliefs or partisan identity—has received comparatively less attention. Yet, discrimination or harassment for political reasons may function in similar ways. Such experiences act as signals of exclusion from public discourse or decision-making, leading individuals to feel marginalized from the political community. This sense of exclusion can cause individuals to question the impartiality, responsiveness, and integrity of institutions.

@@

Importantly, the effects of political discrimination may not be uniform or isolated from the complexities of other political phenomena. In contexts of rising affective polarization, where political identities become emotionally charged and intergroup hostility intensifies (Iyengar et al., 2019; McCoy et al., 2018), perceptions of fairness and legitimacy are increasingly filtered through partisan lenses.

@@

We define political dissonance as the psychological and emotional tension experienced by individuals who perceive a mismatch between their political identity and the dominant political climate or institutional stance (Lelkes, 2016; Federico et al., 2019). Unlike affective polarization, which operates at the societal level, political dissonance is individually experienced and marked by feelings of alienation, frustration, and disillusionment.

Political dissonance is exacerbated in polarized societies where outgroup hostility, moral condemnation, and dehumanization are common (Finkel et al., 2020). This intensifies the emotional toll of political discrimination and heightens perceptions of institutional betrayal (Hetherington \& Rudolph, 2015). In turn, this reduces trust and discourages political participation (Michelson, 2003; Lelkes \& Westwood, 2017).

@@

The second mechanism concerns the politicization of institutional perception—the tendency to evaluate institutions based not on their procedures or performance, but on perceived political alignment. In highly polarized contexts, even neutral institutions are judged through partisan filters (Rogowski \& Sutherland, 2015), leading to a fragmentation of legitimacy. Trust becomes conditional on whether individuals perceive institutions as aligned with their own political views.

@@

This asymmetric trust dynamic reflects the erosion of institutions as shared reference points. Studies show that even traditionally impartial bodies—courts, electoral commissions, public broadcasters—become objects of distrust when seen as favoring one side (Druckman \& Levendusky, 2019; Keefer \& Scartascini, 2022). As Oskooii (2016) notes, perceived exclusion reshapes institutional evaluations, reframing institutions as structurally biased rather than neutral arbiters.

@@

This study uses microdata from general population surveys collected in all 27 European Union Member States as part of the \emph{EUROVOICES} project, conducted by The World Justice Project.\footnote{The World Justice Project \emph{EUROVOICES} was a data collection effort that aimed to evaluate, analyze, and map out national and regional variations in how people in the EU perceive and experience democratic governance, fundamental rights, justice, safety, corruption, transparency, regulatory enforcement, and the business climate (See WJP, 2024).} The dataset provides statistically representative coverage across 110 subnational regions throughout the European Union, offering comprehensive geographic representation of the EU population.

@@

The survey also includes a comprehensive Trust module asking respondents to rate their trust levels in various public institutions as ``a lot", ``some", ``a little", or ``no trust". These institutions include local and national government officials, police officers, prosecutors, public defense attorneys, judges, magistrates, political parties, and Parliament members. We recoded these responses into binary indicators, where ``a lot" and ``some trust" equal 1, and ``a little" and ``no trust" equal 0. To construct a comprehensive measure of political trust, we created a Trust in Political Institutions Index through Logistic Principal Component Analysis (Landgraf \& Lee, 2020), which aggregates individual trust responses across all measured institutions.

@@

The problem is that people who experience political discrimination or harassment might be systematically different from those who do not. For example, individuals who experience such events might be more interested in and informed about politics, or display higher levels of civic participation. If these systematic differences correlate with institutional trust levels, then a naïve comparison between the two groups would be misleading. This is known as the selection problem (Angrist \& Pischke, 2009).

@@

For this study, we implement the matching methodology suggested by Ho et al. (2007) to preprocess our observational data. Our goal is to balance the distribution of covariates between our treatment group (people who experienced D/H events) and control group (people who have not), thereby replicating the randomization achieved in an experimental study (Stuart, 2010).

@@

Importantly, we use matching as a preprocessing method to improve balance between groups, rather than as an imputation method for estimating missing outcomes, as described by Abadie and Imbens (2006, 2016).

We performed nearest neighbor matching\footnote{The matching estimation was performed using the MatchIT R package (Ho et al., 2011).} based on the propensity score, defined as the probability of experiencing a D/H event conditional on a set of observed covariates (Rubin, 1973; Rosenbaum and Rubin, 1983). We estimate this probability using a logistic regression. Each unit from the treatment group is matched with the control group unit that has the closest propensity score (1:1 matching without replacement)\footnote{Matching with replacement means that each control unit can be reused to be matched with any number of treated units if that individual is the closest neighbor to multiple treatment units. Matching without replacement means that each control unit can only be matched with one single treated unit.}. We also establish a region of common support\footnote{The common support region refers to the range of covariate values (or propensity scores) where both treated and control units have sufficient representation.} and discard units that fall outside this region.\footnote{All treatment units fell within the common support region, resulting only in exclusion of individuals from the control group, i.e., people who did not experience political discrimination or harassment.}

@@

This specification is estimated using fixed effects OLS regression\footnote{The estimation of the fixed effects was performed using the fixest R package (Bergé, 2018).}. The value of \( \tau \) can be interpreted as the Average Treatment Effect on the Treated (ATT)\footnote{The Average Treatment Effect (ATE) and the Average Treatment Effect on the Treated (ATT) are two foundational metrics in causal inference. ATE represents the average effect of a treatment across the entire population, whereas ATT focuses only on the average effect among individuals who actually received the treatment.} provided that the conditional ignorability (unconfoundedness) assumption holds (Abadie and Cattaneo, 2018; Greifer and Stuart, 2010).

The unconfoundedness assumption requires that treatment assignment be independent of potential outcomes conditional on the set of observed covariates included. However, even when matching achieves balance in observed covariates, unobserved confounders remain a threat to causal identification. Unobserved variables that relate to both experiencing D/H events and trust in political institutions would violate the unconfoundedness assumption and bias our treatment effect estimates. Since conditional ignorability cannot be directly tested, the literature suggests performing sensitivity analyses to assess the plausibility of this assumption and determine how sensitive our estimated effects are to violations of unconfoundedness (Stuart, 2010).

We use the framework proposed by Cinelli and Hazlett (2020) to assess how fragile are our main results to the possibility of unobserved confounding. This framework allows us to estimate the minimum strength of an unobserved confounder that would be required to nullify our estimated treatment effect. We also assess the robustness of our results to violations of the unconfoundedness assumption by estimating the treatment effect under different levels of unobserved confounding.

We use two measures to test the sensitivity of our results: the robustness value, indicating the minimum confounding strength needed to alter our conclusions, and the partial R2 of the treatment, showing how strongly confounders must be associated with the treatment to completely override our results (Cinelli and Hazlett, 2020).

@@

Appendix \ref{appendix:b} shows balance statistics for the whole sample prior to matching. As we can observe, several covariates display standardized mean differences above the thresholds suggested by the literature—0.1 or, more conservatively, 0.05. Of particular note are the high values for average civic participation, interest in politics, and political alignment with the incumbent party, suggesting that the political profiles of people who have suffered D/H due to their political opinions and those who have not are systematically different. Similarly, age and ethnic composition differ significantly between both groups on average, indicating that the demographic composition of the groups is also systematically different. These insights are validated by the variance ratios and empirical cumulative distribution functions (eCDFs).\footnote{The Std. Mean Diff. is the difference in group means standardized to a single scale for all covariates. The literature suggests values below 0.1—or more conservatively, below 0.05—as indicators of good balance (Ali et al., 2011; Stuart et al., 2013). The Var. Ratio compares the variance of a covariate between the treatment and control groups. Ratios close to 1 indicate good balance (Austin, 2009). Empirical Cumulative Distribution Function Statistics (eCDF) assess balance beyond the mean by comparing the full distribution of each covariate across groups (McCaffrey et al., 2004). Values close to zero indicate good balance.}

@@

Using a matched sample has allowed us to improve the precision of our estimates. However, achieving a balance in covariates using a matching procedure suggests that the treatment and control groups are comparable in ways that are more aligned with the assumptions necessary for causal inference (Branson, 2021). However, balance represents a necessary but not sufficient condition for causality, as unobserved confounders may still bias our estimates. Therefore, we perform a sensitivity analysis to assess the robustness of our results to potential violations of the unconfoundedness assumption.

@@

We assess the robustness of our results to violations of the unconfoundedness assumption using the sensitivity analysis framework of Cinelli and Hazlett (2020). This method estimates the minimum strength of unobserved confounding needed to eliminate our treatment effect and evaluates how results vary under different confounding scenarios. Table \ref{table:5} presents these sensitivity analysis results.

@@

By centering political exclusion, this study extends prior research focused on ethnic and minority-based grievances (Wilkes, 2018; Tyrberg, 2023) to the domain of political and partisan identity, showing that subjective perceptions of injustice based on political beliefs are equally corrosive to democratic legitimacy. This underscores the need to consider how institutional trust is not only eroded by what governments do, but also by how citizens experience their political belonging within the system.

@@

The first is political dissonance, which occurs when individuals feel ideologically misaligned with the dominant political climate. Our findings support the second hypothesis (H2): individuals who both experience political discrimination and feel ideologically out of place exhibit the most pronounced declines in trust. This sense of disconnection intensifies perceptions of exclusion, reinforcing the belief that the political system no longer represents them (McCoy et al., 2018; Finkel et al., 2020).

@@

The second mechanism is politicized perception (Rogowski \& Sutherland, 2015; Druckman \& Levendusky, 2019). In highly polarized contexts, citizens evaluate institutions not based on procedural fairness or performance, but through partisan filters. Our findings support hypothesis three (H3): political discrimination contributes to the reframing of institutions—even neutral ones—as partisan actors.

@@

>>>>

While Europe has historically exhibited lower levels of affective polarization and declining partisan identification, recent studies suggest the presence of partisan hostility and negative feelings driven by factors such as ideological distance, perceived party threats, and regional tensions. Yet, evidence on longitudinal trends is still inconclusive. Scholars continue to emphasize the need for improved measurements and long-term data to determine whether affective polarization is indeed increasing across European democracies \textcite{wagner_affective_2024}.

Despite this uncertainty regarding long-term trends, there is broad agreement in the literature that the rise of far-right populist parties has introduced a distinct affective dimension to European politics—one characterized by heightened hostility toward political out-groups. This emerging divide signals the presence of a particularly polarized and emotionally charged layer of political competition \parencite{reiljan_fear_2020, reiljan_andres_affective_2025}.

Amid this environment, institutional trust is often viewed as a background condition for democratic stability. However, trust plays a more active and central role—it serves as a fundamental mechanism through which political systems gain legitimacy and foster public compliance. A robust body of literature supports the idea that institutional trust enhances democratic resilience, encouraging civic engagement and reinforcing institutional performance \parencite{marien_measuring_2011, devine_stability_2024}.

@@

Institutional trust is often treated as a background condition for democratic stability. Yet, rather than serving as a passive backdrop, it functions as a central mechanism through which political systems earn legitimacy and secure compliance. Democratic institutions rely not only on performance or coercion but on public belief in the fairness of rules, the impartiality of procedures, and the integrity of leaders \parencite{levi_legitimating_2009, citrin_political_2018}.

Trust in institutions reflects more than evaluations of policy outcomes—it signals deeper commitments to institutional principles and the constitutional order \parencite{easton_systems_1965}. Building on this foundation, scholars have shown that institutional trust reinforces civic engagement and perceptions of procedural fairness \parencite{tyler_why_2006}. More recent work highlights its crucial role in promoting democratic resilience, particularly during crises or episodes of institutional contestation \parencite{zmerli_handbook_2017, devine_stability_2024}.

Empirical evidence supports the view that trust strengthens democracy. For instance, \textcite{marien_measuring_2011} finds that institutional trust enables political participation, especially among less resourceful citizens. Other studies show that trust increases when institutions are seen as inclusive and responsive: \textcite{mann_sexual_2022} report higher trust among sexual minorities in countries with equal rights legislation; \textcite{wilkes_immigration_2019} and \textcite{tyrberg_impact_2024} observe similar patterns among ethnic minorities and immigrants.

@@

Supporting this view, \textcite{levi_legitimating_2009} show that perceived government unresponsiveness among disadvantaged groups is associated with lower trust and political efficacy. \textcite{anderson_corruption_2003} and \textcite{oskooii_perceived_2020} similarly find that perceived injustice, whether interpersonal or institutional, generates cognitive and emotional responses that undermine confidence in state institutions.

While this framework has typically focused on ethnic or cultural exclusion \parencite{oskooii_perceived_2020, wilkes_immigration_2019, tyrberg_impact_2024}, political exclusion—based on one's beliefs or partisan identity—has received comparatively less attention. Yet, discrimination or harassment for political reasons may function in similar ways. Such experiences act as signals of exclusion from public discourse or decision-making, leading individuals to feel marginalized from the political community. This sense of exclusion can cause individuals to question the impartiality, responsiveness, and integrity of institutions.

@@

Importantly, the effects of political discrimination may not be uniform or isolated from the complexities of other political phenomena. In contexts of rising affective polarization, where political identities become emotionally charged and intergroup hostility intensifies \parencite{iyengar_origins_2019, mccoy_polarization_2018}, perceptions of fairness and legitimacy are increasingly filtered through partisan lenses.

@@

We define political dissonance as the psychological and emotional tension experienced by individuals who perceive a mismatch between their political identity and the dominant political climate or institutional stance \parencite{lelkes_limits_2017, federico_ideological_2012}. Unlike affective polarization, which operates at the societal level, political dissonance is individually experienced and marked by feelings of alienation, frustration, and disillusionment.

Political dissonance is exacerbated in polarized societies where outgroup hostility, moral condemnation, and dehumanization are common \parencite{iyengar_fear_2015}. This intensifies the emotional toll of political discrimination and heightens perceptions of institutional betrayal \parencite{hetherington_why_2015}. In turn, this reduces trust and discourages political participation \parencite{michelson_corrosive_2003, lelkes_limits_2017}.

@@

The second mechanism concerns the politicization of institutional perception—the tendency to evaluate institutions based not on their procedures or performance, but on perceived political alignment. In highly polarized contexts, even neutral institutions are judged through partisan filters \parencite{rogowski_how_2016, druckman_what_2019}, leading to a fragmentation of legitimacy. Trust becomes conditional on whether individuals perceive institutions as aligned with their own political views.

@@

This asymmetric trust dynamic reflects the erosion of institutions as shared reference points. Studies show that even traditionally impartial bodies—courts, electoral commissions, public broadcasters—become objects of distrust when seen as favoring one side \parencite{keefer_trust_2022}. \textcite{lenz_follow_2012} demonstrates that voters often adjust their perceptions of institutional performance based on partisan cues, even when confronted with neutral or contrary evidence. Similarly, \textcite{nyhan_when_2010} find that political misperceptions persist because citizens interpret information through motivated reasoning, reinforcing distrust in institutions perceived as aligned with opposing political forces. These dynamics are particularly salient among individuals who already feel politically excluded: as \textcite{oskooii_perceived_2020} shows, such experiences of marginalization prompt individuals to reinterpret institutions as structurally biased, rather than as neutral arbiters of the democratic process.

@@

This study uses microdata from general population surveys collected in all 27 European Union Member States as part of the \emph{EUROVOICES} project, conducted by \textcite{the_world_justice_project}.\footnote{The World Justice Project \emph{EUROVOICES} was a data collection effort that aimed to evaluate, analyze, and map out national and regional variations in how people in the EU perceive and experience democratic governance, fundamental rights, justice, safety, corruption, transparency, regulatory enforcement, and the business climate \parencite{the_world_justice_project}.} The dataset provides statistically representative coverage across 110 subnational regions throughout the European Union, offering comprehensive geographic representation of the EU population.

@@

The survey also includes a comprehensive Trust module asking respondents to rate their trust levels in various public institutions as ``a lot", ``some", ``a little", or ``no trust". These institutions include local and national government officials, police officers, prosecutors, public defense attorneys, judges, magistrates, political parties, and Parliament members. We recoded these responses into binary indicators, where ``a lot" and ``some trust" equal 1, and ``a little" and ``no trust" equal 0. To construct a comprehensive measure of political trust, we created a Trust in Political Institutions Index through Logistic Principal Component Analysis \parencite{landgraf_dimensionality_2020}, which aggregates individual trust responses across all measured institutions.

@@

The problem is that people who experience political discrimination or harassment might be systematically different from those who do not. For example, individuals who experience such events might be more interested in and informed about politics, or display higher levels of civic participation. If these systematic differences correlate with institutional trust levels, then a naïve comparison between the two groups would be misleading. This is known as the selection problem \parencite{angrist_mostly_2009}.

@@

For this study, we implement the matching methodology suggested by \textcite{ho_matching_2007} to preprocess our observational data. Our goal is to balance the distribution of covariates between our treatment group (people who experienced D/H events) and control group (people who have not), thereby replicating the randomization achieved in an experimental study \parencite{stuart_matching_2010}.

@@

Importantly, we use matching as a preprocessing method to improve balance between groups, rather than as an imputation method for estimating missing outcomes, as described by \textcite{abadie_large_2006, abadie_matching_2016}.

We performed nearest neighbor matching\footnote{The matching estimation was performed using the MatchIT R package \parencite{ho_matchit_2011}.} based on the propensity score, defined as the probability of experiencing a D/H event conditional on a set of observed covariates \parencite{rubin_matching_1973, rosenbaum_central_1983}. We estimate this probability using a logistic regression. Each unit from the treatment group is matched with the control group unit that has the closest propensity score (1:1 matching without replacement)\footnote{Matching with replacement means that each control unit can be reused to be matched with any number of treated units if that individual is the closest neighbor to multiple treatment units. Matching without replacement means that each control unit can only be matched with one single treated unit.}. We also establish a region of common support\footnote{The common support region refers to the range of covariate values (or propensity scores) where both treated and control units have sufficient representation.} and discard units that fall outside this region.\footnote{All treatment units fell within the common support region, resulting only in exclusion of individuals from the control group, i.e., people who did not experience political discrimination or harassment.}

@@

This specification is estimated using fixed effects OLS regression\footnote{The estimation of the fixed effects was performed using the fixest R package \parencite{berge_efficient_2018}.}. The value of \( \tau \) can be interpreted as the Average Treatment Effect on the Treated (ATT)\footnote{The Average Treatment Effect (ATE) and the Average Treatment Effect on the Treated (ATT) are two foundational metrics in causal inference. ATE represents the average effect of a treatment across the entire population, whereas ATT focuses only on the average effect among individuals who actually received the treatment.} provided that the conditional ignorability (unconfoundedness) assumption holds \parencite{abadie_econometric_2018, greifer_matching_2021}.

The unconfoundedness assumption requires that treatment assignment be independent of potential outcomes conditional on the set of observed covariates included. However, even when matching achieves balance in observed covariates, unobserved confounders remain a threat to causal identification. Unobserved variables that relate to both experiencing D/H events and trust in political institutions would violate the unconfoundedness assumption and bias our treatment effect estimates. Since conditional ignorability cannot be directly tested, the literature suggests performing sensitivity analyses to assess the plausibility of this assumption and determine how sensitive our estimated effects are to violations of unconfoundedness \parencite{stuart_matching_2010}.

We use the framework proposed by \textcite{cinelli_making_2020} to assess how fragile are our main results to the possibility of unobserved confounding. This framework allows us to estimate the minimum strength of an unobserved confounder that would be required to nullify our estimated treatment effect. We also assess the robustness of our results to violations of the unconfoundedness assumption by estimating the treatment effect under different levels of unobserved confounding.

We use two measures to test the sensitivity of our results: the robustness value, indicating the minimum confounding strength needed to alter our conclusions, and the partial R2 of the treatment, showing how strongly confounders must be associated with the treatment to completely override our results \parencite{cinelli_making_2020}.\footnote{The practical implementation of the sensitivity analysis was performed using the sensemakr R package \parencite{cinelli_sensemakr_2024}.}

@@

Appendix \ref{appendix:b} shows balance statistics for the whole sample prior to matching. As we can observe, several covariates display standardized mean differences above the thresholds suggested by the literature—0.1 or, more conservatively, 0.05. Of particular note are the high values for average civic participation, interest in politics, and political alignment with the incumbent party, suggesting that the political profiles of people who have suffered D/H due to their political opinions and those who have not are systematically different. Similarly, age and ethnic composition differ significantly between both groups on average, indicating that the demographic composition of the groups is also systematically different. These insights are validated by the variance ratios and empirical cumulative distribution functions (eCDFs).\footnote{The Std. Mean Diff. is the difference in group means standardized to a single scale for all covariates. The literature suggests values below 0.1—or more conservatively, below 0.05—as indicators of good balance \parencite{ali_propensity_2014, stuart_prognostic_2013}. The Var. Ratio compares the variance of a covariate between the treatment and control groups. Ratios close to 1 indicate good balance \parencite{austin_balance_2009}. Empirical Cumulative Distribution Function Statistics (eCDF) assess balance beyond the mean by comparing the full distribution of each covariate across groups \parencite{mccaffrey_propensity_2004}. Values close to zero indicate good balance.}

@@

Using a matched sample has allowed us to improve the precision of our estimates. However, achieving a balance in covariates using a matching procedure suggests that the treatment and control groups are comparable in ways that are more aligned with the assumptions necessary for causal inference \parencite{branson_randomization_2021}. However, balance represents a necessary but not sufficient condition for causality, as unobserved confounders may still bias our estimates. Therefore, we perform a sensitivity analysis to assess the robustness of our results to potential violations of the unconfoundedness assumption.

@@

We assess the robustness of our results to violations of the unconfoundedness assumption using the sensitivity analysis framework of \textcite{cinelli_making_2020}. This method estimates the minimum strength of unobserved confounding needed to eliminate our treatment effect and evaluates how results vary under different confounding scenarios. Table \ref{table:5} presents these sensitivity analysis results.

@@

By centering political exclusion, this study extends prior research focused on ethnic and minority-based grievances \parencite{wilkes_immigration_2019, tyrberg_impact_2024} to the domain of political and partisan identity, showing that  perceptions of unfair treatment based on political beliefs are equally corrosive to democratic legitimacy. This underscores the need to consider how institutional trust is not only eroded by what governments do, but also by how citizens experience their political belonging within the system.

@@

The first is political dissonance, which occurs when individuals feel ideologically misaligned with the dominant political climate. Our findings support the second hypothesis (H2): individuals who both experience political discrimination and feel ideologically out of place exhibit the most pronounced declines in trust. This sense of disconnection intensifies perceptions of exclusion, reinforcing the belief that the political system no longer represents them \parencite{mccoy_polarization_2018, iyengar_fear_2015}.

@@

The second mechanism is politicized perception \parencite{rogowski_how_2016, druckman_what_2019}. In highly polarized contexts, citizens evaluate institutions not based on procedural fairness or performance, but through partisan filters. Our findings support hypothesis three (H3): political discrimination contributes to the reframing of institutions—even neutral ones—as partisan actors.

@@

% \bibliographystyle{apacite}
% \bibliography{references}
\printbibliography
